Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-07-06 formally.tex
progress on sec 3

\end{subequations}
\subsection{Z-types and Z-expressions}\label{sec:cursors}
\newcommand{\cvert}{{\hspace{1.8px}{\vert}\hspace{1.8px}}}
\begin{figure}[t]
\hspace{-3px}$\arraycolsep=2pt\begin{array}{lllllll}
\mathsf{ZTyp} & \ztau & ::= &
  %\zlsel{\htau} ~\vert~
  \zwsel{\htau} \cvert
  %\zrsel{\htau} ~\vert~
  \tarr{\ztau}{\htau} \cvert
  \tarr{\htau}{\ztau} \\
\mathsf{ZExp} & \zexp & ::= &
  %\zlsel{\hexp} ~\vert~
  \zwsel{\hexp} \cvert
  %\zrsel{\hexp} ~\vert~
  \zexp : \htau \cvert
  \hexp : \ztau \cvert
  \hlam{x}{\zexp} \cvert
  \hap{\zexp}{\hexp} \cvert
  \hap{\hexp}{\zexp} \cvert
  \hadd{\zexp}{\hexp} \cvert
  \hadd{\hexp}{\zexp} \cvert
  \hhole{\zexp}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of Z-types and Z-expressions.}
\label{fig:zexp-syntax}
\end{figure}

@@

The only base cases in these inductive grammars are $\zwsel{\htau}$ and $\zwsel{\hexp}$, which identify the H-type or H-expression that the cursor is on. All of the other forms correspond to the recursive forms in the syntax of H-types and H-expressions, and contain exactly one ``hatted'' subterm that identifies the subtree where the cursor will be found. Any other subterm is ``dotted'', i.e. it is either an H-type or H-expression. Taken together, every Z-type and Z-expression contains exactly one cursor by construction. This can be understood as an instance of Huet's \emph{zipper pattern} \cite{JFP::Huet1997} (which, coincidentally, Huet encountered while implementing a structure editor.)

@@

Similarly, we write $\removeSel{\zexp}$ for the H-expression constructed by erasing the cursor from the Z-expression $\zexp$, i.e. the \emph{cursor erasure} of $\zexp$. The definition of this metafunction is analogous, so we omit it for concision (see supplement.)
% \begin{align*}
% %\removeSel{(\zlsel{\hexp})} & = \hexp\\
% \removeSel{(\zwsel{\hexp})} & = \hexp\\
% %\removeSel{(\zrsel{\hexp})} & = \hexp\\
% \removeSel{(\zexp : \htau)} & = \removeSel{\zexp} : \htau\\
% \removeSel{(\hexp : \ztau)} & = \hexp : \removeSel{\ztau}\\
% \removeSel{(\hlam{x}{\zexp})} & = \hlam{x}{\removeSel{\zexp}}\\ogoo
% \removeSel{(\hap{\zexp}{\hexp})} & = \hap{\removeSel{\zexp}}{\hexp}\\
% \removeSel{(\hap{\hexp}{\zexp})} & = \hap{\hexp}{\removeSel{\zexp}}\\
% \removeSel{(\hadd{\zexp}{\hexp})} & = \hadd{\removeSel{\zexp}}{\hexp}\\
% \removeSel{(\hadd{\hexp}{\zexp})} & = \hadd{\hexp}{\removeSel{\zexp}}\\
% \removeSel{\hhole{\zexp}} &= \hhole{\removeSel{\zexp}}
% \end{align*}

\subsection{Action Semantics}\label{sec:actions}
\begin{figure}[t]
\hspace{-3px}$\arraycolsep=3pt\begin{array}{llcllll}
\mathsf{Action} & \alpha & ::= &
  \aMove{\delta} ~\vert~
  %\aSelect{\delta} ~\vert~
  \aDel ~\vert~
  %\aReplace{\htau} ~\vert~
  %\aReplace{\hexp} ~\vert~
  \aConstruct{\varphi} ~\vert~
  \aFinish\\
\mathsf{Direction} & \delta & ::= &
  \dChild ~\vert~
  \dParent ~\vert~
  \dNext ~\vert~
  \dPrev\\
\mathsf{Shape} & \varphi & ::= &
  \farr ~\vert~
  \fnum \\
& & \vert &
  \fasc ~\vert~
  \fvar{x} ~\vert~
  \flam{x} ~\vert~
  \fap ~\vert~
  \farg ~\vert~
  \fnumlit{n} ~\vert~
  \fplus
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of actions.}
\label{fig:action-syntax}
\vspace{-8px}
\end{figure}

The syntax of \emph{actions}, $\alpha$, some of which involve \emph{directions}, $\delta$, or \emph{shapes}, $\varphi$, is given in Figure \ref{fig:action-syntax}. Actions are performed on Z-types and Z-expressions according to the \emph{action semantics} of Hazelnut, which is organized around three judgements:\todo{fix spacing}
\[\arraycolsep=10pt\begin{array}{ll}
%\textbf{Judgement Form} & \textbf{Description}\\
\performTyp{\ztau}{\alpha}{\ztau'} & \text{Performing $\alpha$ on $\ztau$ produces $\ztau'$}\\
\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'} & \text{Performing $\alpha$ on $\zexp$ when $\removeSel{\zexp}$ synthesizes type $\htau$}\\
& \text{produces $\zexp'$ such that $\removeSel{\zexp'}$ synthesizes type $\htau'$}\\
\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'} & \text{Performing $\alpha$ on $\zexp$ when analyzing $\removeSel{\zexp}$ against $\htau$}\\
& \text{produces $\zexp'$, such that $\removeSel{\zexp'}$ can also be analyzed}\\
& \text{against $\htau$}
\end{array}\]

As suggested by the descriptions above, the action semantics maintains the following \emph{action sensibility} theorem:
\begin{theorem}[Action Sensibility] Both of the following hold:
\label{thrm:actsafe}
\begin{enumerate}
\item If $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
  $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ then
  $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau'}$.
\item If $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
  $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ then
  $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
\end{enumerate}
\end{theorem}
In words, every action leaves the program in a semantically well-defined state. More specifically, the first clause of Theorem \ref{thrm:actsafe} establishes that actions performed on expressions that synthesize a type can only produce expressions that also synthesize some (possibly different) type. The second clause establishes that actions performed on expressions in analytic position (e.g. those under type ascriptions or in argument position, see above) can only produce expressions that can also be analyzed against the expected type.% Non-empty holes allow us to avoid top-down program construction becau but rather can construct fragments of the program inside a hole until ready to ``expose'' them to type analysis.

It is also useful to maintain a \emph{deterministic} action semantics, i.e. every well-defined action should produce a unique Z-type or Z-expression. Formally, this is stated as follows:
\begin{theorem}[Action Determinism] All of the following hold:
\label{thrm:actdet}
\begin{enumerate}
\item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
  $\zexp' = \zexp''$ and $\htau' = \htau''$.
% \item If all of

@@

In order to maintain determinism, we will need to supplement the definition of type compatibility above with a definition for \emph{type incompatibility}, $\tincompat{\htau}{\htau'}$. The key rule establishes that arrow types are incompatible with the $\tnum$ type:
\begin{subequations}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau}{\htau'}
  %   }{
  %     \tincompat{\htau'}{\htau}
  %   }
  % \end{equation}
  \begin{equation}
    \inferrule{ }{
      \tincompat{\tnum}{\tarr{\htau_1}{\htau_2}}
    }
  \end{equation}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_1}{\htau_1'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_2}{\htau_2'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
\end{subequations}
The remaining rules, given in the appendix, establish that type incompatibility is symmetric and covariant.
\subsubsection{Subsumption}

The action semantics includes a subsumption rule much like the one from the underlying semantics of H-expressions:
\begin{equation}
  \inferrule{
    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau'}\\
    \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
    \tcompat{\htau}{\htau''}
  }{
    \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
  }
\end{equation}
In other words, if the expression synthesizes a type, then we defer to the synthetic action performance judgement, as long as it produces an expression that synthesizes a type compatible with the type provided for analysis. It is easy to see that this satisfies Theorem 1 by applying the IH and subsumption.

\subsubsection{Relative Movement} Movement actions change the focus but do not change the underlying H-type or H-expression (so action sensibility is easy to show for these rules as well.)

The rules for relative movement within Z-types are given below and should be self-explanatory:
\begin{subequations}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }{
      \aMove{\dChild}
    }{
      \tarr{\zwsel{\htau_1}}{\htau_2}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \tarr{\zwsel{\htau_1}}{\htau_2}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }
  }
\end{equation}
\begin{equation}\label{rule:move-parent-arr-right}
  \inferrule{ }{
    \performTyp{
      \tarr{{\htau_1}}{\zwsel{\htau_2}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }
  }
\end{equation}
\begin{equation}\label{rule:move-nextSib-arr}
  \inferrule{ }{
    \performTyp{
      \tarr{\zwsel{\htau_1}}{{\htau_2}}
    }{
      \aMove{\dNext}
    }{
      {\tarr{\htau_1}{\zwsel{\htau_2}}}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \tarr{{\htau_1}}{\zwsel{\htau_2}}
    }{
      \aMove{\dPrev}
    }{
      {\tarr{\zwsel{\htau_1}}{{\htau_2}}}
    }
  }
\end{equation}
% \begin{equation}
% \inferrule{
%   \performTyp{
%     \ztau
%   }{
%     \aMove{\delta}
%   }{
%     \ztau'
%   }
% }{
%   \performTyp{
%     \tarr{\ztau}{\htau}
%   }{
%     \aMove{\delta}
%   }{
%     \tarr{\ztau'}{\htau}
%   }
% }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{
%       \ztau
%     }{
%       \aMove{\delta}
%     }{
%       \ztau'
%     }
%   }{
%     \performTyp{
%       \tarr{\htau}{\ztau}
%     }{
%       \aMove{\delta}
%     }{
%       \tarr{\htau}{\ztau}
%     }
%   }
% \end{equation}
\end{subequations}
% The final two rules above recurse into the zipper structure.

The rules for relative movement within Z-expressions are similar. Movement is type-independent, so we defer to an auxiliary judgement for both the analytic and synthetic judgements:\todo{remove prevSib}
\begin{subequations}
\begin{equation}
\inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau}
}
\end{equation}
\begin{equation}
  \inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}
}
\end{equation}
\end{subequations}
For concision, we show only the rules for ascription here:
\begin{subequations}
  \begin{equation}
    \label{r:movefirstchild}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp : \htau}
    }{
      \aMove{\dChild}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveparent}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}\label{rule:move-parent-asc-right}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:movenextsib}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dNext}
    }{
      \hexp : \zwsel{\htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveprevsib}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dPrev}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
\inferrule{
  \performTyp{
    \zexp
  }{
    \aMove{\delta}
  }{
    \zexp'
  }
}{
  \performTyp{
    \zexp : \htau
  }{
    \aMove{\delta}
  }{
    \zexp' : \htau
  }
}
\end{equation}
\begin{equation}
  \inferrule{
    \performTyp{
      \ztau
    }{
      \aMove{\delta}
    }{
      \ztau'
    }
  }{
    \performTyp{
      \hexp : \ztau
    }{
      \aMove{\delta}
    }{
      \hexp : \ztau'
    }
  }
\end{equation}
and this for the example\todo{revise}
\begin{equation}\label{r:movefirstchild-lam}
\inferrule{ }{
  \performMove{
    \zwsel{\hlam{x}{\hexp}}
  }{
    \aMove{\dChild}
  }{
    \hlam{x}{\zwsel{\hexp}}
  }
}
\end{equation}

\begin{equation}\label{r:moveparent-ap2}
  \inferrule{ }{
    \performMove{
      \hap{{\hexp_1}}{\zwsel{\hexp_2}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hap{\hexp_1}{\hexp_2}}
    }
  }
\end{equation}

\begin{equation}\label{r:moveparent-hole}
  \inferrule{ }{
    \performMove{
      \hhole{\zwsel{\hexp}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hhole{\hexp}}
    }
  }
\end{equation}
\end{subequations}

\paragraph{Reachability}

We give the following theorem to demonstrate that the rules for movement do
indeed capture the process of moving the focus to any editable position
within a term. Intuitively, given two zippered expressions that are the
same up to the erasure of focus, its proof gives a list of composable
actions that, when applied to the first term produces the second.

As a consequence of applying Theorem \ref{thrm:actsafe} inductively at
every action in such a list, it is possible consider moving to an ill-typed
term. Therefore the statement here matches the mutually recursive structure
of the bidirectional typing judgements.

\begin{theorem}[Reachability]\todo{why $\htau$ and not $\ztau$?}
\label{thrm:reach}
Both of the following hold:
\begin{enumerate}
  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
   $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
   = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
   ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
   = \zexp'$.

>>>>

\end{subequations}
\subsection{Z-types and Z-expressions}\label{sec:cursors}
\newcommand{\cvert}{{\,{\vert}\,}}
\begin{figure}[t]
\hspace{-3px}$\arraycolsep=2pt\begin{array}{lllllll}
\mathsf{ZTyp} & \ztau & ::= &
  %\zlsel{\htau} ~\vert~
  \zwsel{\htau} \cvert
  %\zrsel{\htau} ~\vert~
  \tarr{\ztau}{\htau} \cvert
  \tarr{\htau}{\ztau} \\
\mathsf{ZExp} & \zexp & ::= &
  %\zlsel{\hexp} ~\vert~
  \zwsel{\hexp} \cvert
  %\zrsel{\hexp} ~\vert~
  \zexp : \htau \cvert
  \hexp : \ztau \cvert
  \hlam{x}{\zexp} \cvert
  \hap{\zexp}{\hexp} \cvert
  \hap{\hexp}{\zexp} \cvert
  \hadd{\zexp}{\hexp} \cvert
  \hadd{\hexp}{\zexp} \cvert
  \hhole{\zexp}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of Z-types and Z-expressions.}
\label{fig:zexp-syntax}
\end{figure}

@@

The only base cases in these inductive grammars are $\zwsel{\htau}$ and $\zwsel{\hexp}$, which identify the H-type or H-expression that the cursor is on. All of the other forms correspond to the recursive forms in the syntax of H-types and H-expressions, and contain exactly one ``hatted'' subterm that identifies the subtree where the focus will be found. Any other sub-term is ``dotted'', i.e. it is either an H-type or H-expression. Taken together, every Z-type and Z-expression contains exactly one focused H-type or H-expression by construction. This can be understood as an instance of Huet's \emph{zipper pattern} \cite{JFP::Huet1997} (which, coincidentally, Huet encountered while implementing a structure editor.)

@@

Similarly, we write $\removeSel{\zexp}$ for the H-expression constructed by erasing the cursor from the Z-expression $\zexp$, i.e. the cursor erasure of $\zexp$. The definition of this metafunction is analogous, so we omit it for concision (see supplement.)
% \begin{align*}
% %\removeSel{(\zlsel{\hexp})} & = \hexp\\
% \removeSel{(\zwsel{\hexp})} & = \hexp\\
% %\removeSel{(\zrsel{\hexp})} & = \hexp\\
% \removeSel{(\zexp : \htau)} & = \removeSel{\zexp} : \htau\\
% \removeSel{(\hexp : \ztau)} & = \hexp : \removeSel{\ztau}\\
% \removeSel{(\hlam{x}{\zexp})} & = \hlam{x}{\removeSel{\zexp}}\\ogoo
% \removeSel{(\hap{\zexp}{\hexp})} & = \hap{\removeSel{\zexp}}{\hexp}\\
% \removeSel{(\hap{\hexp}{\zexp})} & = \hap{\hexp}{\removeSel{\zexp}}\\
% \removeSel{(\hadd{\zexp}{\hexp})} & = \hadd{\removeSel{\zexp}}{\hexp}\\
% \removeSel{(\hadd{\hexp}{\zexp})} & = \hadd{\hexp}{\removeSel{\zexp}}\\
% \removeSel{\hhole{\zexp}} &= \hhole{\removeSel{\zexp}}
% \end{align*}

\subsection{Actions}\label{sec:actions}

We now arrive at the heart of Hazelnut: its \emph{bidirectional action semantics}.  
Figure \ref{fig:action-syntax} defines the syntax of \emph{actions}, $\alpha$, some of which involve \emph{directions}, $\delta$, and \emph{shapes}, $\psi$.

Expression actions are governed by two mutually defined judgements, 1) the \emph{synthetic action judgement}:
\[
\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}
\]
and 2) \emph{the analytic action judgement}:
\[
\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
\]

In some Z-expressions, the cursor is in a type ascription, so we also need a \emph{type action judgement}, pronounced ``performing $\alpha$ on $\ztau$ results in $\ztau'$'':
\[
\performTyp{\ztau}{\alpha}{\ztau'}
\]

Before giving the rules for these judgements, let us state two key metatheorems. These metatheorems deeply inform the design of the rules. Their proofs, which are mechanized in Agda, are by straightforward induction, so the reader is encouraged to think about the relevant cases of these proofs as we present the rules, starting in Sec. \ref{sec:movement}.

% The type action judgement appears as a premise in the rules that handle Z-expressions where the cursor is inside a . 
% \[\begin{array}{ll}
% %\textbf{Judgement Form} & \textbf{Description}\\
% & 
% \end{array}\]
% \[
% \begin{array}{l}
% \fbox{$\performTyp{\ztau}{\alpha}{\ztau'}$} ~~ \text{Performing $\alpha$ on $\ztau$ produces $\ztau'$}\\
% \fbox{$\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$} \\
% \quad \text{Performing $\alpha$ on $\zexp$ when $\removeSel{\zexp}$ synthesizes type $\htau$}\\
% \quad \text{produces $\zexp'$ such that $\removeSel{\zexp'}$ synthesizes type $\htau'$}\\
% \fbox{$\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$} \\
% \quad \text{Performing $\alpha$ on $\zexp$ when analyzing $\removeSel{\zexp}$ against $\htau$}\\
% \quad \text{produces $\zexp'$, such that $\removeSel{\zexp'}$ can also be analyzed}\\
% \quad \text{against $\htau$}
% \end{array}\]
\subsubsection{Sensibility}
The expression action judgements obey the following \emph{action sensibility} theorem:
\begin{theorem}[Action Sensibility]\label{thrm:actsafe} Both of the following hold:\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ 
   then
  $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau'}$.
\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ then
  $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
\end{enumerate}
\end{theorem}
\noindent In other words, if an edit state (i.e. a Z-expression) is statically meaningful, i.e. its cursor erasure is well-typed, then performing an action on it leaves the resulting edit state statically meaningful. In particular, the first clause of Theorem \ref{thrm:actsafe} establishes that when an action is performed on an edit state whose cursor erasure synthesizes an H-type, the result is an edit state whose cursor erasure also synthesizes some (possibly different) H-type. The second clause establishes that when an action is performed using the analytic action judgement on an edit state whose cursor erasure analyzes against some H-type, the result is a Z-expression whose cursor erasure also analyzes against the same H-type. % Non-empty holes allow us to avoid top-down program construction becau but rather can construct fragments of the program inside a hole until ready to ``expose'' them to type analysis.

The top-level sequence of actions in Figures \ref{fig:first-example} and \ref{fig:second-example} begin with an empty hole, so the derivation of each edit step must conclude with the synthetic action judgement. The analytic action judgement and the type action judgement appear as premises in the derivations of some of these steps.

\begin{figure}[t]
\hspace{-3px}$\arraycolsep=3pt\begin{array}{llcllll}
\mathsf{Action} & \alpha & ::= &
  \aMove{\delta} ~\vert~
  %\aSelect{\delta} ~\vert~
  \aDel ~\vert~
  %\aReplace{\htau} ~\vert~
  %\aReplace{\hexp} ~\vert~
  \aConstruct{\psi} ~\vert~
  \aFinish\\
\mathsf{Dir} & \delta & ::= &
  \dChild ~\vert~
  \dParent ~\vert~
  \dNext \\ % ~\vert~
  % \dPrev\\
\mathsf{Shape} & \psi & ::= &
  \farr ~\vert~
  \fnum \\
& & \vert &
  \fasc ~\vert~
  \fvar{x} ~\vert~
  \flam{x} ~\vert~
  \fap ~\vert~
  \farg ~\vert~
  \fnumlit{n} ~\vert~
  \fplus\\
& & \vert & 
  {\color{gray}\fnehole}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of actions.}
\label{fig:action-syntax}
\end{figure}

\subsubsection{Determinism}
Another useful property that we maintain is \emph{action determinism}, i.e. that performing an action should produce a unique result. Formally, this is established as follows:
\begin{theorem}[Action Determinism] All of the following hold:
\label{thrm:actdet}
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
  $\zexp' = \zexp''$ and $\htau' = \htau''$.
% \item If all of

@@

In order to maintain determinism, we will need to supplement our definition of type consistency from Figure \ref{fig:type-consistency} with a definition of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$. One can define this notion either direrctly as the metatheoretic negation of type consistency, or as a separate inductively defined judgement. The supplement does the latter. The key rule establishes that arrow types are inconsistent with $\tnum$:
% \begin{subequations}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau}{\htau'}
  %   }{
  %     \tincompat{\htau'}{\htau}
  %   }
  % \end{equation}
  \begin{equation*}
    \inferrule{ }{
      \tincompat{\tnum}{\tarr{\htau_1}{\htau_2}}
    }
  \end{equation*}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_1}{\htau_1'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_2}{\htau_2'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
% \end{subequations}
% The remaining rules, given in the appendix, establish that type incompatibility is symmetric.
The mechanization proves that this judgemental definition of type inconsistency is indeed the negation of type consistency.\todo{big if true!!!}

\subsubsection{Relative Movement}\label{sec:movement} 
The rules below define relative movement within Z-types. They should be self-explanatory:
\begin{subequations}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }{
      \aMove{\dChild}
    }{
      \tarr{\zwsel{\htau_1}}{\htau_2}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \tarr{\zwsel{\htau_1}}{\htau_2}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }
  }
\end{equation}
\begin{equation}\label{rule:move-parent-arr-right}
  \inferrule{ }{
    \performTyp{
      \tarr{{\htau_1}}{\zwsel{\htau_2}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }
  }
\end{equation}
\begin{equation}\label{rule:move-nextSib-arr}
  \inferrule{ }{
    \performTyp{
      \tarr{\zwsel{\htau_1}}{{\htau_2}}
    }{
      \aMove{\dNext}
    }{
      {\tarr{\htau_1}{\zwsel{\htau_2}}}
    }
  }
\end{equation}
% \begin{equation}
%   \inferrule{ }{
%     \performTyp{
%       \tarr{{\htau_1}}{\zwsel{\htau_2}}
%     }{
%       \aMove{\dPrev}
%     }{
%       {\tarr{\zwsel{\htau_1}}{{\htau_2}}}
%     }
%   }
% \end{equation}
% \begin{equation}
% \inferrule{
%   \performTyp{
%     \ztau
%   }{
%     \aMove{\delta}
%   }{
%     \ztau'
%   }
% }{
%   \performTyp{
%     \tarr{\ztau}{\htau}
%   }{
%     \aMove{\delta}
%   }{
%     \tarr{\ztau'}{\htau}
%   }
% }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{
%       \ztau
%     }{
%       \aMove{\delta}
%     }{
%       \ztau'
%     }
%   }{
%     \performTyp{
%       \tarr{\htau}{\ztau}
%     }{
%       \aMove{\delta}
%     }{
%       \tarr{\htau}{\ztau}
%     }
%   }
% \end{equation}
\end{subequations}
Two more rules are needed to recurse into the zipper structure. We define these zipper rules in an action-independent manner in Sec. \ref{sec:zipper-cases}. 
% The final two rules above recurse into the zipper structure.

The rules for relative movement within Z-expressions are similarly straightforward. Movement is type-independent, so we defer to an auxiliary expression movement judgement in both the analytic and synthetic case:\todo{remove prevSib}
\begin{subequations}
\begin{equation}
\inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau}
}
\end{equation}
\begin{equation}
  \inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}
}
\end{equation}
\end{subequations}
% \begin{figure}
% \judgbox{\performMove{\zexp}{\aMove{\delta}}{\zexp'}}{(Additional movement-action rules:)}
% \begin{displaymath}
% \begin{array}{@{}rcl}
%   %\multicolumn{3}{l}{\textbf{Movement Actions for Sum Type Forms}:}
%   %\\
%   \TABperformMove
%       {\zwsel{\hexp : \htau}}
%       {\aMove{\dChild}}
%       {      \zwsel{\hexp} : \htau}
%   \\
%   \TABperformMove
%       {
%       \zwsel{\hexp} : \htau
%     }{
%       \aMove{\dParent}
%     }{
%       \zwsel{\hexp : \htau}
%     }
%   \\
%   \TABperformMove
%     {
%       \hexp : \zwsel{\htau}
%     }{
%       \aMove{\dParent}
%     }{
%       \zwsel{\hexp : \htau}
%     }
%   \\
%   \TABperformMove
%       {
%       \zwsel{\hexp} : \htau
%     }{
%       \aMove{\dNext}
%     }{
%       \hexp : \zwsel{\htau}
%     }
%   \\[2mm]
%   \TABperformMove
%       {      {\tsum{{\htau_1}}{{\zwsel{\htau_2}}}}}
%       {\aMove{\dParent}}
%       {\zwsel{\tsum{{\htau_1}}{\htau_2}}}
%   \\[2mm]
%   \TABperformMove
%       {\zwsel{\hinj{i}{\hexp}}}
%       {\aMove{\dChild}}
%       {\hinj{i}{\zwsel{\hexp}}}      
%   \\
%   \TABperformMove
%       {\hinj{i}{\zwsel{\hexp}}}      
%       {\aMove{\dParent}}
%       {\zwsel{\hinj{i}{\hexp}}}
%   \\[2mm]
%   \TABperformMove
%       {\zwsel{\hcase{\hexp}{x}{\hexp_1}{y}{\hexp_2}}}
%       {\aMove{\dChild}}
%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%       {\aMove{\dNext}}
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%       {\aMove{\dNext}}
%       {      {\hcase{{\hexp}}{x}{{\hexp_1}}{y}{\zwsel{\hexp_2}}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{{\hexp_1}}{y}{\zwsel{\hexp_2}}}}
%       {\aMove{\dPrev}}
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%       {\aMove{\dPrev}}
%       {      {\hcase{{\zwsel{\hexp}}}{x}{{\hexp_1}}{y}{{\hexp_2}}}}
%   \\
%   \TABperformMove
%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%       {\aMove{\dParent}}
%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%       {\aMove{\dParent}}
%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\hexp_1}{y}{\zwsel{\hexp_2}}}}
%       {\aMove{\dParent}}
%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
% \end{array}
% \end{displaymath}
% \caption{Movement action semantics for sum type forms.}
% \end{figure}
The expression movement judgement is defined as follows.
